Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-01-22 paper.tex
Edit intro and discussion; fix numerical section.

The source localization problem is one that appears in many fields where it is
of interest to find the position of a source using an array of sensors.

Our principal motivation comes from an application in brain activity
measurement, where one wishes to use modalities such as electroencephalography
(EEG) and magnetoencephalography (MEG) to record activity inside the
brain~\cite{Baillet2001Electromagnetic}.  For example, current EEG systems use
anywhere between 10 and 250 sensors (electrodes) placed on the scalp, to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are frequently modeled as current dipoles. For a single
activated dipole within the brain (which is modeled as being localized to a
point), the electrodes effectively sense a diffuse (blurred-out) representation
of the point source, as a result of the spatial low-pass filtering effect of
the different layers between the brain surface and scalp surface (cerebrospinal
fluid, skull and skin).

Recent developments~\cite{Grover2016Information} have suggested that source
localization can benefit greatly from an increase in sensor density. This
conclusion is based primarily on a Nyquist rate analysis, and on an
information-theoretic bound (which are possibly very loose) on the accuracy of
source localization algorithms. In deriving this bound, however, the scalp
potential was never sampled; equivalently, it was assumed that an
\emph{infinite} number of sensors was available to sample the EEG signal at
every point on the scalp. Hence, the dependence of the bound on the number of
sensors is lost.

Earlier work by Mosher et al.~\cite{Mosher1993Error}, which seeks to address
this issue by deriving Cramer-Rao lower bounds, is also not completely
satisfactory because these bounds do not apply to biased estimators. Commonly
used source localization algorithms (such as the Minimum Norm
Estimate~\cite{Hamalainen1994Interpreting}) on the other hand, are \emph{known}
to be biased~\cite{Lin2006Assessing}.

This paper seeks to address the shortcomings of earlier methods by deriving
minimax lower bounds for the source localization problem, which apply to
\emph{all} estimators (biased \emph{and} unbiased), and which can give a
scaling in terms of the number of sensors used (for a fixed sensor placement
configuration). Computing the lower bound for the ``real brain model'' (where
the complete geometry of the brain surface is accounted for) is rendered very
difficult since analytical solutions for the EEG signal, given the brain
activity, do not exist. ``Spherical head
models''~\cite{Nunez2006Electric,Grover2016Information} \emph{do} have
analytical solutions, but are still not easily amenable to analytical lower
bounds because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}, and because the solution is expressed in
terms of special functions. We therefore restrict our analysis to a source
localization problem on a one-dimensional domain (to be described in detail in
Section~\ref{sec:source-localization}).

@@

Hence, we believe that this paper is the first to give lower bounds on source
localization error (measured in an $\ell_2$-distance sense) in estimating the
location of a one-dimensional point source, which is observed by sensors
through a diffusive medium (treated as a low-pass filter), and corrupted by
additive Gaussian noise. We describe the problem setting in detail in
Section~\ref{sec:source-localization}, the minimax techniques and our main
result in Section~\ref{sec:minimax-lower-bounds} and a few straightforward
extensions in Section~\ref{sec:extensions}. Finally, we conclude with a
discussion on implications for sensor placement and future work in
Section~\ref{sec:discussion}.

@@

We start by giving a detailed description of the one-dimensional setting of the
source localization problem. Recall that, in this problem, we're trying to
determine the location of a point source by observing it through a diffusive
medium using a sensor array.

@@

\subsection{Towards a more realistic sensor model}

The bound in Theorem~\ref{thm:main-lb} indicates that the error in source
localization goes to zero, as the number of sensors goes to infinity. This
observation is distinct from the lower bound given
in~\cite{Grover2016Fundamental}, where even with an infinite number of sensors,
the lower bound is strictly greater than zero. This indicates that the bound
in section~\ref{sec:lecam-lb} is loose for large numbers of sensors.

The looseness lies in a poor sensor model: we assume that an arbitrarily large
number of sensors can be squeezed into a finite amount of space, and that these
sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is an impractical expectation. In reality, sensors have a
certain width, and the signal measured by them is the integral of the
continuous-space signal within that width (this is similar to a sample-and-hold
circuit, which integrates a continuous-\emph{time} signal over a short period).
If the width of a sensor is decreased, then effectively, its signal-to-noise
ratio reduces \textcolor{red}{<Need to add reference or explanation>} (this is
intuitively evident in the sample-and-hold analogy: to compute the signal
value, the integral must be divided by the time interval over which the
integration is performed; this division amplifies noise as the time interval
gets smaller).

We performed a preliminary study to understand the effect of viewing sensors
as integrators. The bounding technique used in section~\ref{sec:lecam-lb} no
longer easily admits a solution which gives a bound for all values of $m$. We
can compute the bound numerically, however, using exact expressions for error
probability in a binary hypothesis test. For gaussian distributions of equal
variance, the expression is a Q-function in the distance between their means:
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_1 - \v\mu_2}}{2\sigma}\bigg)
\end{equation}
\textcolor{red}{Is this definitely correct? i.e.\ the best test is a linear
separator? I think this is coming from Neyman-Pearson. Should I cite?}
The minimax estimation risk is lower bounded by the risk in hypothesis testing,
so
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) = \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma}\bigg).
\end{equation}
This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
separated by a distance of $2\delta$. As the $\delta$ increases, it becomes
easier to distinguish between the two hypotheses, so $P_e$ falls. When
numerically evaluating the bound, therefore, we have to \emph{manually}
trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
hypotheses to distinguish between. To make the simulation easier, we can relax
the process of finding the least discernable hypothesis pair by fixing
$\theta_0$. Then, we need only optimize over $\delta$, making the process more
tractable. This comes at the cost of loosening the lower bound.

@@

\section{Extensions}
\label{sec:extensions}

\subsection{From a circular (periodic) domain to a linear (aperiodic) domain}

So far, we've relied on the presence of a circular domain for to make the
arguments in the proof. However, the most important requirement was that
\emph{all} parts of the impulse response should get \emph{sensed} uniformly.
So, if $\theta$ were restricted to the region $[0, S)$ (now on a linear,
non-periodic domain), and if sensors were available to sense the signal from
$[-w/2, S{+}w/2)$ (for an impulse response with restricted support as described
in section~\ref{sec:signal-model}), then the proof for sensor noise would still
fall through, albeit with slightly different constants.

@@

\subsection{From a one-dimensional domain to a $d$-dimensional domain}

Another simple extension takes our problem from a one-dimensional setting to a
$d$-dimensional setting. Here, we would have $\v\theta \in [0, S)^d$, and the
$\ell_2$ loss function, $\Phiorho(\v\theta, \vhat\theta) = \norm{\v\theta -
\vhat\theta}^2$. Then, using Assouad's lemma~\cite{Tsybakov2009Introduction} in
the manner described in~\cite{Duchi2015Information}, we can split the minimax
risk along each of the $d$ dimensions, and bound each separately using Le Cam's
method.

@@

The lower bound given by Theorem~\ref{thm:main-lb} goes to zero as the number
of sensors, $m$, goes to infinity, \emph{even if} the total number of trials
$n$ is kept \emph{fixed}. This would mean that a high sensor density can
overcome the effect of noise in a single trial. The reason this happens is that
we assume sensors are affected by i.i.d. noise, no matter what sampling density
we use. This is an increasingly poor assumption as sensor density increases,
however; we expect noise to become correlated if sensors are extremely close
(due to inductive/capacitive coupling, for example). A more realistic lower
bound will show the benefit of increasing sensor density up to a point only,
beyond which it will have to scale alongside the number of trials.

The lower bounding \emph{technique} gives certain inisights on the issue of
sensor placement. Suppose we wanted to achieve a higher resolution in a
restricted interval of the parameter space, i.e.\ $\theta \in [a, b] \subset
\Theta$, such that $\abs{b - a} \ll w$. Then, it might be of greatest benefit
to place sensors \emph{not} precisely over the region, but rather in the region
corresponding to the \emph{largest change} in the impulse response. Indeed, it
is possible to derive an  approximate lower bound using Taylor's approximation,
which shows that the KL-divergence between distributions is maximized (and
hence error minimized) at regions where the slope of $x(s;\theta)$ is largest.

Another point on sensor placement bears repeating: the bounds derived in this
paper apply to only a specific sensor configuration, i.e., the uniform one.
This would intuitively seem to be an appropriate choice in the minimax setting,
however, since any non-uniform placement will have one or more points which is
far from all sensors. The ``maximization'' over $\theta \in \Theta$ would then
select such a point, where the value of the impulse response so produced at the
sensors is smallest, so that SNR is minimized.

>>>>

The source localization problem is one that arises in many fields in different
forms. Our principal motivation comes from Electroencephalography (EEG), a
non-invasive brain measurement modality which uses sensors (electrodes) placed
on the scalp to sense electric potentials produced by neuronal activity within
the brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are frequently modeled as current dipoles, and it is
often of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as being localized to a point), the
electrodes effectively sample a diffuse (blurred) representation of the point
source, as a result of the spatial low-pass filtering effect of the different
layers between the brain surface and scalp surface (cerebrospinal fluid, skull
and skin).

Recent developments~\cite{Grover2016Information} have suggested that source
localization can benefit greatly from an increase in sensor density. This
conclusion is based primarily on a Nyquist rate analysis, and on an
information-theoretic bound (first derived in~\cite{Grover2016Fundamental}) on
the accuracy of source localization algorithms. This information-theoretic
bound, however, was derived for the unsampled, continuous-space potential on
the scalp. In other words, it was a result asymptotic in the number of sensors.
The claim in~\cite{Grover2016Information} is therefore not fully substantiated
by this analysis, since it does not provide a lower bound showing improvement
with increasing sensor density.

Earlier work by Mosher et al.~\cite{Mosher1993Error} also sought to derive
lower bounds on source localization accuracy by deriving Cramer-Rao lower
bounds. They also have some preliminary numerical results showing improvement
with increased numbers of sensors. However, this result is also not completely
satisfactory because Cramer-Rao bounds do not apply to biased estimators.
Commonly used source localization algorithms (such as the Minimum Norm
Estimate~\cite{Hamalainen1994Interpreting}) on the other hand, are \emph{known}
to be biased~\cite{Lin2006Assessing}.

This paper seeks to address the shortcomings of earlier methods by deriving
lower bounds on the minimax risk for the source localization problem, which
apply to \emph{all} estimators (biased \emph{and} unbiased), and which can give
a scaling in terms of the number of sensors used (for a fixed sensor placement
configuration). Computing the lower bound for the ``real brain model'' (where
the complete geometry of the brain surface is accounted for) is rendered very
difficult since analytical solutions for the EEG signal, given the brain
activity, do not exist. Previous lower bounds are based on ``spherical head
models''~\cite{Nunez2006Electric,Grover2016Information} which \emph{have}
analytical solutions. But these are still not easily amenable to analytical
lower bounds because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}, and because the solution is expressed in
terms of special functions. We therefore restrict our analysis to a source
localization problem on a one-dimensional domain (to be described in detail in
Section~\ref{sec:source-localization}).

@@

Hence, we believe that this paper is the first to give lower bounds on source
localization error (measured in an $\ell_2$-distance sense) in estimating the
location of a one-dimensional point source, which is observed by sensors
through a diffusive medium (treated as a low-pass filter), and corrupted by
additive Gaussian noise. We describe the problem setting in detail in
Section~\ref{sec:source-localization}, the minimax techniques and our main
results in Section~\ref{sec:minimax-lower-bounds}. Finally, we conclude in
Section~\ref{sec:discussion} with a discussion on how our results can be
extended to other simple settings, and on implications for sensor placement.

@@

We start by giving a detailed description of the one-dimensional setting of the
source localization problem.

@@

\subsection{Towards a more \textcolor{blue}{physical} sensor model}

The bound in Theorem~\ref{thm:main-lb} indicates that the error in source
localization goes to zero, as the number of sensors goes to infinity. This
observation is significantly different from the information-theoretic lower
bound given in~\cite{Grover2016Fundamental}, where even with an infinite number
of sensors, the lower bound is strictly greater than zero. This indicates that
the bound in section~\ref{sec:lecam-lb} is loose for large numbers of sensors.

The looseness lies in a poor sensor model: we assume that an arbitrarily large
number of sensors can be squeezed into a finite amount of space, and that these
sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is an impractical expectation. In reality, sensors have a
certain width, and the signal measured by them is the integral of the
continuous-space signal within that width (this is similar to a sample-and-hold
circuit, which integrates a continuous-\emph{time} signal over a short period).
If the width of a sensor is decreased, then effectively, its signal-to-noise
ratio reduces (this is intuitively evident in the sample-and-hold analogy: to
compute the signal value, the integral must be divided by the time interval
over which the integration is performed; this division amplifies noise as the
time interval gets smaller).

We performed a preliminary study to understand the effect of viewing sensors
as integrators. In the new sensor model, sensors have a finite width, which
decreases as the number of sensors increases, in order to allow their number to
grow unbounded. Concretely, the $k$-th sensor which is located at $s = kS/m$
extends from $(k{-}\frac{1}{2})S/m$ to $(k{+}\frac{1}{2})S/m$. The signal
sensed by this sensor is the $k$-th element of $\v x(\theta)$, given by
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds
\end{equation}
\textcolor{red}{Also need to talk about noise here - sensor obs. are still iid
because we assume wss noise, and integrating over distinct regions of this
process retains iid nature}
The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a solution which gives a bound for all values of $m$. We can compute the
bound numerically, however, using exact expressions for error probability in a
binary hypothesis test. For gaussian distributions of equal variance, the best
possible error probability (over all tests) is a Q-function in the distance
between their means:
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma}\bigg)
\end{equation}
\textcolor{red}{Is this definitely correct? i.e.\ the best test is a linear
separator? I think this is coming from Neyman-Pearson. Should I cite?}
The minimax estimation risk is lower bounded by the risk in hypothesis testing,
hence
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) = \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma}\bigg).
\end{equation}
This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
separated by a distance of $2\delta$. As the $\delta$ increases, it becomes
easier to distinguish between the two hypotheses, so $P_e$ falls. When
numerically evaluating the bound, therefore, we have to \emph{manually}
trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
hypotheses to distinguish between. To make the simulation easier, we can relax
the process of finding the least discernable hypothesis pair by fixing
$\theta_0$. Then, we need only optimize over $\delta$, making the process more
tractable. This comes at the cost of loosening the lower bound.

@@

%\section{Extensions}
%\label{sec:extensions}

%\subsection{From a circular (periodic) domain to a linear (aperiodic) domain}

%So far, we've relied on the presence of a circular domain for to make the
%arguments in the proof. However, the most important requirement was that
%\emph{all} parts of the impulse response should get \emph{sensed} uniformly.
%So, if $\theta$ were restricted to the region $[0, S)$ (now on a linear,
%non-periodic domain), and if sensors were available to sense the signal from
%$[-w/2, S{+}w/2)$ (for an impulse response with restricted support as described
%in section~\ref{sec:signal-model}), then the proof for sensor noise would still
%fall through, albeit with slightly different constants.

@@

%\subsection{From a one-dimensional domain to a $d$-dimensional domain}

%Another simple extension takes our problem from a one-dimensional setting to a
%$d$-dimensional setting. Here, we would have $\v\theta \in [0, S)^d$, and the
%$\ell_2$ loss function, $\Phiorho(\v\theta, \vhat\theta) = \norm{\v\theta -
%\vhat\theta}^2$. Then, using Assouad's lemma~\cite{Tsybakov2009Introduction} in
%the manner described in~\cite{Duchi2015Information}, we can split the minimax
%risk along each of the $d$ dimensions, and bound each separately using Le Cam's
%method.

@@

While the lower bound presented in section~\ref{sec:lecam-lb} is non-physical
for a large number of sensors, it still gives some important insights. For
instance, if we assume that sensors have a constant width $c$, then even if we
view sensors as integrators, the bound tells us something about the scaling of
error with number of sensors, up to $m = \floor*{\frac{S}{c}}$. Beyond this
point, averaging noise by increasing the number of trials may help one get to a
target mean squared error in location. An analytical derivation of the
trade-off between $m$ and $n$ in achieving a certain target MSE is relegated to
future work.

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a one-dimensional linear domain
(of fixed length) and to a $d$-dimensional domain (circular or linear) are
fairly straightforward, under certain conditions. For a 1D linear domain, we
require that sensors capture the entirety of the impulse response even at the
edges of the domain. Hence, sensors must be placed uniformly in $[-w/2,
S{+}w/2]$. For extending to $d$-dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent 1D problems.

The lower bounding \emph{technique} itself gives certain inisights on the issue
of sensor placement. Suppose we wanted to achieve a higher resolution in a
restricted interval of the parameter space, i.e.\ $\theta \in [a, b] \subset
\Theta$, such that $\abs{b - a} \ll w$. Then, it might be of greatest benefit
to place sensors \emph{not} precisely over $[a, b]$, but rather in the region
\emph{surrounding} $[a, b]$, corresponding to the \emph{largest change} in the
impulse response. Indeed, it is possible to derive an approximate lower bound
using Taylor's approximation, which shows that the KL-divergence between
distributions is maximized (and hence error minimized) at regions where the
slope of $x(s;\theta)$ is largest.

Another point on sensor placement bears repeating: the bounds derived in this
paper apply to only a specific sensor configuration, i.e., the uniform one.
This would intuitively seem to be an appropriate choice in the minimax setting,
however, since any non-uniform placement will have one or more points which is
far from all sensors. The ``maximization'' over $\theta \in \Theta$ would then
select a $\theta$ for which the value of the impulse response at the sensors is
smallest, so that SNR is minimized.
